One-Particle Density Matrix Polarization Susceptibility Tensors

### 2018-08-02 papers/1_dmatpol/paper-r3.tex
Revision

In the limit of a weak and spatially uniform electric field,
the polarization\hyp{}induced density matrix change is given by the follwoing Taylor expansion:
%
\begin{equation}\label{e:dD-Taylor}
 \delta D_{\alpha\beta} = \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}}  \cdot {\bf F} 
   + \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} : {\bf F} \otimes {\bf F}
   + \ldots 
\end{equation}
%
In the above expression, the density matrix susceptibilities are simply equal to the derivatives of the density matrix
with respect to electric field (up to a multiplicative constant).
In a general case, i.e., when electric field is non\hyp{}uniform, the development of a first\hyp{}principles theory
that is capable of accurately capturing the electric field\hyp{}dependence of the one\hyp{}particle density
becomes extremely difficult. This is not only because of the very complicated (and generally unknown) solution
to the electronic correlation problem, but also because
a working model has to accurately predict all the $n^2$ elements of the density matrix
where $n$ is the size of basis set. Even relatively small errors
in a particular matrix element will accumulate when performing two\hyp{}
and four\hyp{}index contractions,
and in turn, affect the evaluated properties with non\hyp{}negligible errors.
Single\hyp{}centered expansions, like the one from Eq.~\eqref{e:dD-Taylor}, although valid for 
an arbitrary level of theory,
are not sufficient enough to accurately describe the local polarization
of electron density distribution when electric fields are highly non\hyp{}uniform in space.
Due to all of these reasons, we adopt the following approach: 
we first develop a qualitative first\hyp{}principles distributed\hyp{}site model at the Hartree\hyp{}Fock
level of theory, and then generalize it to the arbitrary level using the functional form that was found 
by considering HF level,
but treating the density matrix susceptibility tensors as cetrain functions of adjustable parameters further on.

@@

\begin{acknowledgments}
This project is carried out under POLONEZ programme which has received funding from the European Union's
Horizon~2020 research and innovation programme under the Marie Skłodowska-Curie grant agreement 
No.~665778. This project is funded by National Science Centre, Poland 
(grant~no. 2016/23/P/ST4/01720) within the POLONEZ 3 fellowship.
Many thanks to Robert W.~G{\'o}ra and Robert Zale{\'s}ny 
for reading and commenting on the manuscript.
\end{acknowledgments}

@@

Multipole integrals depend on the origin with respect to which they are evaluated.
However, the induced dipole moments defined in Eq.~\eqref{e:dmu-4-exact.linear-approximation} 
(or equivalently in Eq.~\eqref{e:dmu-l-vector})
have to be origin independent. 
Let us start with noticing that
%
\begin{equation}
 \left[ \widetilde{\mathbb{M}} \right]_{i\alpha} ({\bf r}_Q) 
 = \left[ \widetilde{\mathbb{M}} \right]_{i\alpha} ({\bf 0}) - \left[ {\bf C}^\dagger \right]_{i\alpha} {\bf r}_Q  \;.
\end{equation}
%
Further we have
%
\begin{equation}
 \left[ \widetilde{\mathbb{K}} \right]_{i\alpha} ({\bf r}_Q) 
 = \left[ \widetilde{\mathbb{K}} \right]_{i\alpha} ({\bf 0}) - \left[ {\bf C}^\dagger \cdot {\bf D} \right]_{i\alpha} {\bf r}_Q \;.
\end{equation}
%
But ${\bf C}^\dagger \cdot {\bf D}={\bf C}^\dagger \cdot {\bf C} \cdot {\bf C}^\dagger={\bf C}^\dagger$ in orthogonal
AO basis. This implies that
%
\begin{multline}
   \left[ \widetilde{\mathbb{L}} \right]_{i\alpha} ({\bf r}_Q) 
 = \left[ \widetilde{\mathbb{M}} \right]_{i\alpha} ({\bf r}_Q) 
 - \left[ \widetilde{\mathbb{K}} \right]_{i\alpha} ({\bf r}_Q) \\
 = \left[ \widetilde{\mathbb{M}} \right]_{i\alpha} ({\bf 0})   
 - \left[ \widetilde{\mathbb{K}} \right]_{i\alpha} ({\bf 0})
 = \left[ \widetilde{\mathbb{L}} \right]_{i\alpha} ({\bf 0}) \;.
\end{multline}
%
Therefore, it is proved that the polarization\hyp{}induced distributed dipole moments 
defined in Eq.~\eqref{e:dmu-4-exact.linear-approximation} 
are origin independent.
Thus, one can compute dipole integrals with respect to any origin and resulting
susceptibilities ${\bf B}_{\alpha\beta}^{(i;1)}$ from Eq.~\eqref{e:susceptibility-B} will be uniquely defined.

>>>>

In the limit of a weak and spatially uniform electric field,
the polarization\hyp{}induced density matrix change is given by the following Taylor expansion:
%
\begin{equation}\label{e:dD-Taylor}
 \delta D_{\alpha\beta} = \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}}  \cdot {\bf F} 
   + \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} : {\bf F} \otimes {\bf F}
   + \ldots 
\end{equation}
%
In the above expression, the density matrix susceptibilities are simply equal to the derivatives of the density matrix
with respect to electric field (up to a multiplicative constant).
In a general case, i.e., when electric field is non\hyp{}uniform, the development of a first\hyp{}principles theory
that is capable of accurately capturing the electric field\hyp{}dependence of the one\hyp{}particle density
becomes extremely difficult. This is not only because of the very complicated (and generally unknown) solution
to the electronic correlation problem, but also because
a working model has to accurately predict all the $n^2$ elements of the density matrix
where $n$ is the size of basis set. Even relatively small errors
in a particular matrix element will accumulate when performing two\hyp{}
and four\hyp{}index contractions,
and in turn, affect the evaluated properties with non\hyp{}negligible errors.
Single\hyp{}centered expansions, like the one from Eq.~\eqref{e:dD-Taylor}, although valid for 
an arbitrary level of theory,
are not sufficient enough to accurately describe the local polarization
of electron density distribution when electric fields are highly non\hyp{}uniform in space.
Due to all of these reasons, we adopt the following approach: 
we first develop a qualitative first\hyp{}principles distributed\hyp{}site model at the Hartree\hyp{}Fock
level of theory, and then generalize it to the arbitrary level using the functional form that was found 
by considering HF level,
but treating the density matrix susceptibility tensors as cetrain functions of adjustable parameters further on.

@@

\begin{acknowledgments}
This project is carried out under POLONEZ programme which has received funding from the European Union's
Horizon~2020 research and innovation programme under the Marie Skłodowska-Curie grant agreement 
No.~665778. This project is funded by National Science Centre, Poland 
(grant~no. 2016/23/P/ST4/01720) within the POLONEZ 3 fellowship.
%Many thanks to Robert W.~G{\'o}ra and Robert Zale{\'s}ny 
%for reading and commenting on the manuscript.
\end{acknowledgments}

@@

Multipole integrals depend on the origin with respect to which they are evaluated.
However, the induced dipole moments defined in Eq.~\eqref{e:dmu-4-exact.linear-approximation} 
(or equivalently in Eq.~\eqref{e:dmu-l-vector})
have to be origin independent. Note that
%
\begin{equation}
 \left[ \widetilde{\mathbb{M}} \right]_{i\alpha} ({\bf r}_Q) 
 = \left[ \widetilde{\mathbb{M}} \right]_{i\alpha} ({\bf 0}) - \left[ {\bf C}^\dagger \right]_{i\alpha} {\bf r}_Q  
\end{equation}
%
and
%
\begin{equation}
 \left[ \widetilde{\mathbb{K}} \right]_{i\alpha} ({\bf r}_Q) 
 = \left[ \widetilde{\mathbb{K}} \right]_{i\alpha} ({\bf 0}) - \left[ {\bf C}^\dagger \cdot {\bf D} \right]_{i\alpha} {\bf r}_Q \;.
\end{equation}
%
But ${\bf C}^\dagger \cdot {\bf D}={\bf C}^\dagger \cdot {\bf C} \cdot {\bf C}^\dagger={\bf C}^\dagger$ in orthogonal
AO basis which implies that
%
\begin{multline}
   \left[ \widetilde{\mathbb{L}} \right]_{i\alpha} ({\bf r}_Q) 
 = \left[ \widetilde{\mathbb{M}} \right]_{i\alpha} ({\bf r}_Q) 
 - \left[ \widetilde{\mathbb{K}} \right]_{i\alpha} ({\bf r}_Q) \\
 = \left[ \widetilde{\mathbb{M}} \right]_{i\alpha} ({\bf 0})   
 - \left[ \widetilde{\mathbb{K}} \right]_{i\alpha} ({\bf 0})
 = \left[ \widetilde{\mathbb{L}} \right]_{i\alpha} ({\bf 0}) \;.
\end{multline}
%
Therefore, it is proved that the polarization\hyp{}induced distributed dipole moments 
defined in Eq.~\eqref{e:dmu-4-exact.linear-approximation} are origin independent.
Thus, one can compute dipole integrals with respect to any origin and resulting
susceptibilities ${\bf B}_{\alpha\beta}^{(i;1)}$ from Eq.~\eqref{e:susceptibility-B} will be uniquely defined.
